Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-05-22 r-journal/manuscript.tex
update with Ken's comments

As seen throughout this paper, we display the network obtained from
cohesion using a force-directed graph drawing algorithm and emphasize
the strong ties (colored by connected component). We refer to the
connected components of the network of strong ties as ``clusters.'' Note
that to qualify as a cluster in this definition, one may not have any
strong ties with those outside the cluster, and thus the existence of
disjoint groups is a strong signal for separation. Here, clusters are
identified without additional user inputs nor optimization criteria. If
one wishes to further break the community graph into groups, one may use
community detection methods for networks (such as spectral clustering or
the Louvain algorithm). Though only briefly considered here, one may
also use the collection of strong ties in place of (weighted) k-nearest
neighbors in settings such as classification and smoothing. Overall, the
structural information obtained from local depth, cohesion and community
graphs can provide a holistic perspective on the data which does not
require the use of distributional assumptions, optimization criteria nor
additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depths (PaLD) probabilities, implementing clustering analyses, and
creating data visualizations to represent the clusters. This paper will
describe how to use the package as well as walk through two examples.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Helper functions to organize data into the correct format, into
  distance matrices and then cohesion matrices
\item
  Functions that convert a cohesion matrix into a variety of useful
  formats, including partitioned local depths, clusters, and graphs
\item
  Plotting functions
\end{enumerate}

@@

For demonstration purposes, below is a sample data frame with two
variables, \texttt{x1} and \texttt{x2}. The methods put forth here work
on data frames with higher dimensions, as described in the
\textbf{Examples} section; we are simply choosing a small data frame
here for demonstration purposes.

\begin{Schunk}
\begin{Sinput}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4, 14),
  x2 = c(5, 4, 13, 7, 6, 10)
)
rownames(df) <- c("A", "B", "C", "D", "E", "F")
\end{Sinput}
\end{Schunk}

The first step needed to calculate the partitioned local depths is to
construct a \emph{distance matrix}. If the data are already in this
form, the user can skip to the next step. The \texttt{dist()} function
converts an input data frame into a distance matrix, as demonstrated
below.

@@

This will create a \texttt{dist} object. If converted to a matrix, this
will be a \(n\times n\) distance matrix, where \(n\) corresponds to the
number of observations in the original data frame, in this example
\(n = 6\).

This dist object, or equivalently a distance matrix, can then be passed
to the \texttt{cohesion\_matrix()} function in order to calculate the
pairwise cohesion values. Cohesion is an interpretable probability that
reflects the strength of alignment of two points. Again, if the user
begins with a distance matrix, they can skip the first step and simply
input the distance matrix into this function.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
cohesion_matrix(d)
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1666667 0.0000000 0.0000000 0.1666667 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.1000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.1000000 0.0000000 0.1000000
#> D 0.0000000 0.0000000 0.1000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.1000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.1000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1666667 0.0000000 0.0000000 0.1666667 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.1000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.1000000 0.0000000 0.1000000
#> D 0.0000000 0.0000000 0.1000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.1000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.1000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

The \emph{cohesion matrix} output by the \texttt{cohesion\_matrix()} is
the main input for the majority of the remaining functions.

\hypertarget{functions-that-convert-a-cohesion-matrix-into-useful-formats}{%
\subsubsection{Functions that convert a cohesion matrix into useful
formats}\label{functions-that-convert-a-cohesion-matrix-into-useful-formats}}

@@

To calculate the \emph{clusters} that each point will fall into, we can
use the \texttt{community\_clusters()} function. This will output a data
frame with two columns, the first will correspond to the \texttt{point},
as identified by the row name of the original input data frame,
\texttt{df}, the second will identify the \texttt{cluster} that each
point belongs to.

@@

\begin{Schunk}
\begin{Sinput}
local_depths(cohesion)
\end{Sinput}
\begin{Soutput}
#>         A         B         C         D         E         F 
#> 0.5666667 0.4666667 0.4333333 0.5333333 0.4666667 0.5333333
\end{Soutput}
\end{Schunk}

@@

The \texttt{strong\_threshold()} function will calculate the cohesion
threshold for strong ties. This is equal to half the average of the
diagonal cohesion matrix. This is a threshold that may be used to
distinguish between strong and weak ties.

\begin{Schunk}
\begin{Sinput}
strong_threshold(cohesion)
\end{Sinput}
\begin{Soutput}
#> [1] 0.1222222
\end{Soutput}
\end{Schunk}

In this case, the threshold is \texttt{0.12}.

@@

The function \texttt{cohesion\_strong()} will update the cohesion matrix
to set all weak ties to zero (via the \texttt{strong\_threshold()}
function). Optionally, the matrix will also be symmetrized, with the
default parameter \texttt{symmetric\ =\ TRUE}.

\begin{Schunk}
\begin{Sinput}
cohesion_strong(cohesion)
\end{Sinput}
\begin{Soutput}
#>           A         B         C         D         E         F
#> A 0.2333333 0.1333333 0.0000000 0.0000000 0.1333333 0.0000000
#> B 0.1333333 0.2333333 0.0000000 0.0000000 0.0000000 0.0000000
#> C 0.0000000 0.0000000 0.2333333 0.0000000 0.0000000 0.0000000
#> D 0.0000000 0.0000000 0.0000000 0.2666667 0.0000000 0.1666667
#> E 0.1333333 0.0000000 0.0000000 0.0000000 0.2333333 0.0000000
#> F 0.0000000 0.0000000 0.0000000 0.1666667 0.0000000 0.2666667
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

Finally, the \texttt{community\_graphs()} function takes the cohesion
matrix and creates \CRANpkg{igraph} objects, graphs that describes the
relationship between the points. This function will output a list of
three objects:

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH b3faedc UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from b3faedc (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

Here we see that there are three connected components, points \texttt{A}
and \texttt{B}, and \texttt{A} and \texttt{E} which form the first
cluster, and points \texttt{D} and \texttt{F} which form another.

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
a square plot so as to not distort the distances. When using the
\CRANpkg{ggplot2} package for this visualization, you can use the
\texttt{aspect.ratio\ \ =\ 1} argument in the \texttt{theme()} function.
If using the \texttt{plot()} function included in the base library, you
can set the plot area to be a square by using \texttt{par(pty\ =\ "s")}
prior to calling the \texttt{plot()} function, as shown below.

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df)) + 
  theme(aspect.ratio = 1)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `df`]{Visualize the points from data frame `df`}\label{fig:fig1}
\end{figure}
\end{Schunk}

@@

The \texttt{layout} argument allows the user to pass a matrix to dictate
the layout of the graph. For example, if we wanted the graph to match
the visualization displayed in Figure \ref{fig:fig1}, we can pass
\texttt{as.matrix(df)}, or a matrix of the data frame \texttt{df} to the
\texttt{layout} argument (Figure \ref{fig:fig3}. Notice in the code
below, we set \texttt{par(pty\ =\ "s")} in order to ensure that we have
a square plotting region. Additionally, this
\texttt{plot\_community\_graphs()} function will also permit parameters
that can be passed to the \texttt{plot.igraph()} function. For example,
we can pass arguments to the \texttt{plot.igraph} function via the
\texttt{...}, for example to increase the vertex size and change the
vertex label color, we can specify \texttt{vertex.size\ =\ 15} and
\texttt{vertex.label.color\ =\ "white"}.

@@

plot_community_graphs(cohesion, 
                      layout = as.matrix(df),
                      vertex.size = 15,
                      vertex.label.color = "white")
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

@@

The first example if from a subset of tissue gene expression data from
\citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. A \texttt{dist} object was created
using this data set and is included the \CRANpkg{pald} package in an
object called \texttt{tissue\_dist}.

@@

We can also display the relationships between tissue samples using the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}). For
clarity of the display, we show how to remove the labels using
\texttt{show\_labels\ =\ FALSE}. We will instead color by the labels by
passing these to the \texttt{vertex.color} parameter through the
\texttt{...} to the \texttt{plot.igraph} function. Similarly, we can add
a legend using the \texttt{legend()} function, as you would for a
\CRANpkg{igraph} visualization. Additionally, we use the
\texttt{edge\_width\_factor} and \texttt{emph\_strong} arguments to
adjust the width of the lines between and within PaLD clusters.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{fig5} \caption[PaLD clustering of tissue data]{PaLD clustering of tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

We can then use the \texttt{neighbors()} function from the
\CRANpkg{igraph} package to extract the neighbors in this graph. For
example, if we wanted to extract all neighbors where the language is
``French'', we would run the following.

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 46d19b2:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

The \CRANpkg{pald} includes three randomly generated data frames
corresponding to plots from \citet{berenhaut2022social}:

@@

We will demonstrate how we can compare PaLD to two clustering methods:
\emph{k}-means and hierarchical clustering. The code below calculates
the cohesion matrix (\texttt{exdata\_cohesion}) as well as the clusters
via PaLD (\texttt{exdata\_pald}), \emph{k}-means
(\texttt{exdata\_kmeans}) and hierarchical clustering
(\texttt{exdata\_hclust}).

@@

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)"
)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to *k*-means and hierarchical clustering with k = 8.}\label{fig:fig5}
\end{figure}
\end{Schunk}

>>>>

As seen throughout this paper, we display the network obtained from
cohesion using a force-directed graph drawing algorithm and emphasize
the strong ties (colored by connected component). We refer to the
connected components of the network of strong ties as community
``clusters.'' Note that to qualify as a cluster in this definition, one
may not have any strong ties with those outside the cluster, and thus
the existence of disjoint groups is a strong signal for separation.
Here, clusters are identified without additional user inputs nor
optimization criteria. If one wishes to further break the community
graph into groups, one may use community detection methods for networks
(such as spectral clustering or the Louvain algorithm). Though only
briefly considered here, one may also use the collection of strong ties
in place of (weighted) k-nearest neighbors in settings such as
classification and smoothing. Overall, the structural information
obtained from local depth, cohesion and community graphs can provide a
holistic perspective on the data which does not require the use of
distributional assumptions, optimization criteria nor additional user
inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depths (PaLD) probabilities, implementing clustering analyses, and
creating data visualizations to display community structure. This paper
will describe how to use the package as well as walk through two
examples.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A function for computing the cohesion matrix
\item
  Functions for extracting useful information from the cohesion matrix,
  such as local depths, neighbors, clusters, and graph objects
\item
  Plotting functions for community graphs
\end{enumerate}

@@

The input for the Partitioned Local Depths (PaLD) is a distance matrix
or \texttt{dist} object. Note that the collection of input distances (or
dissimilarities) does not need to satisfy the triangle inequality.

For demonstration purposes, we will show how one can compute a distance
matrix from an input data frame with, say, two variables \texttt{x1} and
\texttt{x2}. The input data may be of any dimension. The methods put
forth here work on data frames with higher dimensions, as described in
the \textbf{Examples} section; we are simply choosing a small data frame
here for demonstration purposes.

\begin{Schunk}
\begin{Sinput}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 8, 16, 4, 14),
  x2 = c(5, 4, 10, 8, 4, 10)
)
rownames(df) <- c("A", "B", "C", "D", "E", "F")
\end{Sinput}
\end{Schunk}

The \texttt{dist()} function converts an input data frame into a
distance matrix, as demonstrated below. If the data are already provided
as a distance matrix (or \texttt{dist} object), the user can skip to the
next step.

@@

The function above creates a \texttt{dist} object. If converted to a
matrix, this will be a \(n\times n\) distance matrix, where \(n\)
corresponds to the number of observations in the original data frame, in
this example \(n = 6\).

This \texttt{dist} object, or a distance matrix, can then be passed to
the \texttt{cohesion\_matrix()} function in order to calculate the
pairwise cohesion values.

Cohesion is an interpretable probability that reflects the strength of
alignment of two points within local regions. It captures aspects of the
relative positioning of points and accounts for varying density across
the space.

\begin{Schunk}
\begin{Sinput}
d <- dist(df)
cohesion_matrix(d)
\end{Sinput}
\begin{Soutput}
#>            A          B          C         D          E         F
#> A 0.25000000 0.18333333 0.06666667 0.0000000 0.18333333 0.0000000
#> B 0.14000000 0.24000000 0.05000000 0.0000000 0.10666667 0.0000000
#> C 0.07333333 0.07333333 0.20333333 0.0000000 0.03333333 0.0800000
#> D 0.00000000 0.00000000 0.00000000 0.2333333 0.00000000 0.1333333
#> E 0.14000000 0.10666667 0.03333333 0.0000000 0.24000000 0.0000000
#> F 0.00000000 0.00000000 0.05000000 0.1400000 0.00000000 0.2400000
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
df |>
  dist() |>
  cohesion_matrix()
\end{Sinput}
\begin{Soutput}
#>            A          B          C         D          E         F
#> A 0.25000000 0.18333333 0.06666667 0.0000000 0.18333333 0.0000000
#> B 0.14000000 0.24000000 0.05000000 0.0000000 0.10666667 0.0000000
#> C 0.07333333 0.07333333 0.20333333 0.0000000 0.03333333 0.0800000
#> D 0.00000000 0.00000000 0.00000000 0.2333333 0.00000000 0.1333333
#> E 0.14000000 0.10666667 0.03333333 0.0000000 0.24000000 0.0000000
#> F 0.00000000 0.00000000 0.05000000 0.1400000 0.00000000 0.2400000
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

The \emph{cohesion matrix} output by the \texttt{cohesion\_matrix()}
function is the main input for the majority of the remaining functions.

\hypertarget{functions-for-extracting-information-from-the-cohesion-matrix}{%
\subsubsection{Functions for extracting information from the cohesion
matrix}\label{functions-for-extracting-information-from-the-cohesion-matrix}}

@@

@@

\begin{Schunk}
\begin{Sinput}
local_depths(cohesion)
\end{Sinput}
\begin{Soutput}
#>         A         B         C         D         E         F 
#> 0.6833333 0.5366667 0.4633333 0.3666667 0.5200000 0.4300000
\end{Soutput}
\end{Schunk}

@@

The \texttt{strong\_threshold()} function will calculate the cohesion
threshold for strong ties. This is equal to half the average of the
diagonal cohesion matrix.\citep{berenhaut2022social} This is a threshold
that may be used to distinguish between strong and weak ties.

\begin{Schunk}
\begin{Sinput}
strong_threshold(cohesion)
\end{Sinput}
\begin{Soutput}
#> [1] 0.1172222
\end{Soutput}
\end{Schunk}

In this case, the threshold is \texttt{0.117}.

The function \texttt{cohesion\_strong()} will update the cohesion matrix
to set all weak ties to zero (via the \texttt{strong\_threshold()}
function). Optionally, the matrix will also be symmetrized, using the
entry-wise (parallel) minimum of the cohesion matrix and its transpose,
with the default parameter \texttt{symmetric\ =\ TRUE}.

\begin{Schunk}
\begin{Sinput}
cohesion_strong(cohesion)
\end{Sinput}
\begin{Soutput}
#>      A    B         C         D    E         F
#> A 0.25 0.14 0.0000000 0.0000000 0.14 0.0000000
#> B 0.14 0.24 0.0000000 0.0000000 0.00 0.0000000
#> C 0.00 0.00 0.2033333 0.0000000 0.00 0.0000000
#> D 0.00 0.00 0.0000000 0.2333333 0.00 0.1333333
#> E 0.14 0.00 0.0000000 0.0000000 0.24 0.0000000
#> F 0.00 0.00 0.0000000 0.1333333 0.00 0.2400000
#> attr(,"class")
#> [1] "cohesion_matrix" "matrix"          "array"
\end{Soutput}
\end{Schunk}

The \texttt{community\_graphs()} function takes the cohesion matrix and
creates \CRANpkg{igraph} objects, graphs that describe the relationship
between the points. This function will output a list of three objects:

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH 8eef810 UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from 8eef810 (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

Here we see that there are three connected components, ties \texttt{A-B}
and and \texttt{A-E} which form the first cluster, and the tie
\texttt{D-F} which forms another.

@@

The ``clusters'' identified by PaLD are the connected components of the
graph of strong ties, \texttt{G\_strong}. To directly calculate them, we
can use the \texttt{community\_clusters()} function. This will output a
data frame with two columns, the first will correspond to the
\texttt{point}, as identified by the row name of the original input data
frame, \texttt{df}, the second will identify the \texttt{cluster} that
each point belongs to.

@@

The final category of function is functions for data visualization. We
can begin by visualizing the points in data frame \texttt{df} (Figure
\ref{fig:fig1}). When visualizing these points, it is important to have
the aspect ratio of the x and y axes equal to 1 so as to not distort the
distances. When using the \CRANpkg{ggplot2} package for this
visualization, you can use the \texttt{coord\_fixed(ratio\ =\ 1)}
function. If using the \texttt{plot()} function included in the base
library, you can use the \texttt{asp\ =\ 1} argument.

\begin{Schunk}
\begin{Sinput}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df)) + 
  coord_fixed(ratio = 1) + 
  xlim(c(4, 16)) + 
  ylim(c(4, 16))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig1-1} \caption[Visualize the points from data frame `df`]{Visualize the points from data frame `df`}\label{fig:fig1}
\end{figure}
\end{Schunk}

@@

The \texttt{layout} argument allows the user to pass a matrix to dictate
the 2-dimensional layout of the graph. For example, if we wanted the
graph to match the visualization displayed in Figure \ref{fig:fig1}, we
can pass \texttt{as.matrix(df)}, or a matrix of the data frame
\texttt{df} to the \texttt{layout} argument (Figure \ref{fig:fig3}).
Additionally, this \texttt{plot\_community\_graphs()} function will also
permit parameters that can be passed to the \texttt{plot.igraph()}
function. For example, we can pass arguments to the \texttt{plot.igraph}
function via \texttt{...}, for example to increase the vertex size and
change the vertex label color, we can specify
\texttt{vertex.size\ =\ 100} and
\texttt{vertex.label.color\ =\ "white"}. Additionally, to allow axes, we
use \texttt{axes\ =\ TRUE} and to put them back on the original scale we
set \texttt{rescale\ =\ FALSE}, resetting the axis limits using
\texttt{xlim} and \texttt{ylim}. The \texttt{par(pty\ =\ "s")} function
forces the subsequent plot to be square.

@@

plot_community_graphs(cohesion, 
                      layout = as.matrix(df),
                      vertex.size = 100,
                      vertex.label.color = "white",
                      axes = TRUE,
                      rescale = FALSE,
                      asp = 1,
                      xlim = c(4, 16),
                      ylim = c(4, 16))
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig3-1} \caption[PaLD graph displaying the relationship between the points in data frame `df`, matching the original layout in Figure 1]{PaLD graph displaying the relationship between the points in data frame `df`, matching the original layout in Figure 1}\label{fig:fig3}
\end{figure}
\end{Schunk}

@@

The first example is from a subset of tissue gene expression data from
\citet{zilliox2007gene}, \citet{mccall2011gene}, and
\citet{mccall2014gene}, obtained from the \textbf{tissuesGeneExpression}
bioconductor package \citep{tissue}. A \texttt{dist} object was created
using this data set and is included the \CRANpkg{pald} package in an
object called \texttt{tissue\_dist}.

@@

We can also display the relationships between tissue samples using the
\texttt{plot\_community\_graphs()} function (Figure \ref{fig:fig4}). For
clarity of the display, we show how to remove the labels using
\texttt{show\_labels\ =\ FALSE}. We will instead color by the labels by
passing these to the \texttt{vertex.color} parameter through the
\texttt{...} to the \texttt{plot.igraph} function. Similarly, we can add
a legend using the \texttt{legend()} function, as you would for an
\CRANpkg{igraph} visualization. Additionally, we use the
\texttt{edge\_width\_factor} and \texttt{emph\_strong} arguments to
adjust the width of the lines between and within PaLD clusters.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{fig5} \caption[Community cluster network for the tissue data]{Community cluster network for the tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.}\label{fig:fig4}
\end{figure}
\end{Schunk}

@@

We can then use the \texttt{neighbors()} function from the
\CRANpkg{igraph} package to extract the strong neighbors in this graph.
For example, if we wanted to extract all neighbors where the language is
``French'', we would run the following.

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 9c17296:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

@@

The \CRANpkg{pald} package includes three randomly generated data frames
corresponding to plots from \citet{berenhaut2022social}:

@@

We will demonstrate how we can compare PaLD to two clustering methods:
\emph{k}-means and hierarchical clustering. The code below calculates
the cohesion matrix (\texttt{exdata\_cohesion}) as well as the clusters
via PaLD (\texttt{exdata\_pald}), \emph{k}-means
(\texttt{exdata\_kmeans}) and hierarchical clustering using complete
linkage (\texttt{exdata\_hclust}).

@@

\begin{Schunk}
\begin{Sinput}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)",
  asp = 1
)
\end{Sinput}
\begin{figure}
\includegraphics{manuscript_files/figure-latex/fig5-1} \caption[PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al]{PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to k-means and hierarchical clustering with k = 8.}\label{fig:fig5}
\end{figure}
\end{Schunk}
